---
title: "Michelson Speed-of-light Measurements"
author: "Esther Aduamah"
date: 2025-02-06
output: 
  github_document:
    toc: true
prerequisites:
  - e-data02-derive
editor_options: 
  markdown: 
    wrap: 72
---

*Purpose*: When studying physical problems, there is an important
distinction between *error* and *uncertainty*. The primary purpose of
this challenge is to dip our toes into these factors by analyzing a real
dataset.

*Reading*: [Experimental Determination of the Velocity of
Light](https://play.google.com/books/reader?id=343nAAAAMAAJ&hl=en&pg=GBS.PA115)
(Optional)

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics
define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|------------------------|------------------------|------------------------|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and
supporting files (`report_files/` folder) when you are done! Then submit
a link to Canvas. **Your Challenge submission is not complete without
all files uploaded to GitHub.**

```{r message=FALSE, warning=FALSE}
# Libraries
library(tidyverse)
library(googlesheets4)

url <- "https://docs.google.com/spreadsheets/d/1av_SXn4j0-4Rk0mQFik3LLr-uf0YdA06i3ugE6n-Zdo/edit?usp=sharing"

# Parameters
LIGHTSPEED_VACUUM    <- 299792.458 # Exact speed of light in a vacuum (km / s)
LIGHTSPEED_MICHELSON <- 299944.00  # Michelson's speed estimate (km / s)
LIGHTSPEED_PM        <- 51         # Michelson error estimate (km / s)
```

*Background*: In 1879 Albert Michelson led an experimental campaign to
measure the speed of light. His approach was a development upon the
method of Foucault[3], and resulted in a new estimate of
$v_0 = 299944 \pm 51$ kilometers per second (in a vacuum). This is very
close to the modern *exact* value of `r LIGHTSPEED_VACUUM`. In this
challenge, you will analyze Michelson's original data, and explore some
of the factors associated with his experiment.

I've already copied Michelson's data from his 1880 publication; the code
chunk below will load these data from a public googlesheet.

*Aside*: The speed of light is *exact* (there is **zero error** in the
value `LIGHTSPEED_VACUUM`) because the meter is actually
[*defined*](https://en.wikipedia.org/wiki/Metre#Speed_of_light_definition)
in terms of the speed of light!

```{r read-sheet}
## Note: No need to edit this chunk!
gs4_deauth()
ss <- gs4_get(url)
df_michelson <-
  read_sheet(ss) %>%
  select(Date, Distinctness, Temp, Velocity) %>%
  mutate(Distinctness = as_factor(Distinctness))

df_michelson %>% glimpse()
```

*Data dictionary*:

-   `Date`: Date of measurement
-   `Distinctness`: Distinctness of measured images: 3 = good, 2 = fair,
    1 = poor
-   `Temp`: Ambient temperature (Fahrenheit)
-   `Velocity`: Measured speed of light (km / s)

### **q1** Re-create the following table (from Michelson (1880), pg. 139) using `df_michelson` and `dplyr`. Note that your values *will not* match those of Michelson *exactly*; why might this be?

| Distinctness | n   | MeanVelocity |
|--------------|-----|--------------|
| 3            | 46  | 299860       |
| 2            | 39  | 299860       |
| 1            | 15  | 299810       |

```{r q1-task}
## TODO: Compute summaries
library(dplyr)

df_q1 <- df_michelson %>%
  group_by(Distinctness) %>%
  summarize(
    n = n(),
    mean_velocity = round(mean(Velocity), 0)
  ) %>%
  arrange(desc(Distinctness)) 

# Display the table
df_q1 %>%
  knitr::kable()
```

**Observations**: - Write your observations here! - The distinctness
refers to the clarity/quality of the images that Michelson observed in
his experiments. The higher distinctness, the clearer measurements and
vice versa. There were graded on three levels of measurments. n is the
number of trials. Mean velocity is the average speed of light. The two
mean velocities with distinctness 3 & 2 are practically the same. THis
table allows us to see how the clairty of the measurements affected the
sppeed of light estimates. Why might your table differ from
Michelson’s? - There could be slight differences in how the measurements
were recorded. He could have used a different way to sort his data and
the set may not have perfectly matched his.

The `Velocity` values in the dataset are the speed of light *in air*;
Michelson introduced a couple of adjustments to estimate the speed of
light in a vacuum. In total, he added $+92$ km/s to his mean estimate
for `VelocityVacuum` (from Michelson (1880), pg. 141). While the
following isn't fully rigorous ($+92$ km/s is based on the mean
temperature), we'll simply apply this correction to all the observations
in the dataset.

### **q2** Create a new variable `VelocityVacuum` with the $+92$ km/s adjustment to `Velocity`. Assign this new dataframe to `df_q2`.

```{r q2-task}
## TODO: Adjust the data, assign to df_q2
df_q2 <- df_michelson %>%
  mutate(VelocityVacuum = Velocity + 92) 

df_q2
```

As part of his study, Michelson assessed the various potential sources
of error, and provided his best-guess for the error in his
speed-of-light estimate. These values are provided in
`LIGHTSPEED_MICHELSON`---his nominal estimate---and
`LIGHTSPEED_PM`---plus/minus bounds on his estimate. Put differently,
Michelson believed the true value of the speed-of-light probably lay
between `LIGHTSPEED_MICHELSON - LIGHTSPEED_PM` and
`LIGHTSPEED_MICHELSON + LIGHTSPEED_PM`.

Let's introduce some terminology:[2]

-   **Error** is the difference between a true value and an estimate of
    that value; for instance `LIGHTSPEED_VACUUM - LIGHTSPEED_MICHELSON`.
-   **Uncertainty** is an analyst's *assessment* of the error.

Since a "true" value is often not known in practice, one generally does
not know the error. The best they can do is quantify their degree of
uncertainty. We will learn some means of quantifying uncertainty in this
class, but for many real problems uncertainty includes some amount of
human judgment.[2]

### **q3** Compare Michelson's speed of light estimate against the modern speed of light value. Is Michelson's estimate of the error (his uncertainty) greater or less than the true error?

```{r q3-task}
## TODO: Compare Michelson's estimate and error against the true value
## Your code here!

michelson_error <- LIGHTSPEED_VACUUM - LIGHTSPEED_MICHELSON

michelson_uncertainty <- LIGHTSPEED_PM

michelson_error

# Quantitative comparison
error_ratio <- abs(michelson_error) / michelson_uncertainty
error_difference <- abs(michelson_error) - michelson_uncertainty

list(
  michelson_error = michelson_error,
  michelson_uncertainty = michelson_uncertainty,
  error_ratio = error_ratio,
  error_difference = error_difference
)
```

**Observations**: -

Is Michelson’s estimate of the error (his uncertainty) greater or less
than the true error? -

This indicates that Michelson underestimated the uncertainty or
potential error in his experiment. His estimated uncertainty was smaller
than the actual difference between his measured speed of light and the
true value. -\
Make a quantitative comparison between Michelson’s uncertainty and his
error. -

Michelson’s true error (the actual deviation of his measurement from the
true speed of light) is larger than his uncertainty (the range he
believed the true value might lie in). Michelson’s True Error: -151.542
km/s and Michelson’s Uncertainty: 51 km/s as per the formula: True Error
= Lightspeed_Vacuum - Lightspeed_Michelson.

The following plot shows all of Michelson's data as a [control
chart](https://en.wikipedia.org/wiki/Control_chart); this sort of plot
is common in manufacturing, where it is used to help determine if a
manufacturing process is under [statistical
control](https://en.wikipedia.org/wiki/Statistical_process_control).
Each dot is one of Michelson's measurements, and the grey line connects
the mean taken for each day. The same plot also shows simulated data
using a probability model. We'll get into statistics later in the
course; for now, let's focus on understanding what real and simulated
data tend to look like.

### **q4** Inspect the following plot with the `Real` Michelson data and `Simulated` data from a probability model. Document the similarities and differences between the data under *observe* below.

```{r q4-cf-real-simulated}
## Note: No need to edit this chunk!
## Calibrate simulated data
v_mean <-
  df_q2 %>%
  summarize(m = mean(VelocityVacuum)) %>%
  pull(m)
v_sd <-
  df_q2 %>%
  summarize(s = sd(VelocityVacuum)) %>%
  pull(s)

## Visualize
set.seed(101)
df_q2 %>%
  mutate(Simulated = rnorm(n(), mean = v_mean, sd = v_sd)) %>%
  rename(Real = VelocityVacuum) %>%
  pivot_longer(
    cols = c(Simulated, Real),
    names_to = "source",
    values_to = "velocity"
  ) %>%

  ggplot(aes(Date, velocity)) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON,
    linetype = "dotted"
  ) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON - LIGHTSPEED_PM,
    linetype = "dashed"
  ) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON + LIGHTSPEED_PM,
    linetype = "dashed"
  ) +

  geom_line(
    data = . %>%
      group_by(Date, source) %>%
      summarize(velocity_mean = mean(velocity)),
    mapping = aes(y = velocity_mean),
    color = "grey50"
  ) +
  geom_point(
    mapping = aes(y = velocity),
    size = 0.8
  ) +

  facet_grid(source~.) +
  theme_minimal() +
  labs(
    x = "Date of Measurement (1879)",
    y = "Velocity (in Vacuum)"
  )
```

**Observations**:

Similarities:

-   Both data sets show similar overall ranges of velocity values

-   the daily means (gray lines) follow similar patterns of variation in
    both data sets

-   Both show points scattered around their respective means

Differences

-   The real data shows more consistent daily patterns- measurements
    from the same day tend to cluster together

-   The simulated data appears more randomly scattered without the daily
    grouping patterns

### **q5** You have access to a few other variables. Construct a **at least three** visualizations of `VelocityVacuum` against these other factors. Are there other patterns in the data that might help explain the difference between Michelson's estimate and `LIGHTSPEED_VACUUM`?

```{r}
# Visualization 1: Velocity vs Temperature
ggplot(df_q2, aes(Temp, VelocityVacuum, color = Distinctness)) +
  geom_point(alpha = 0.5, size = 4) +
  geom_smooth(method = "lm", se = FALSE, aes(group = Distinctness)) +
  scale_color_manual(values = c("1" = "red", "2" = "green", "3" = "blue")) +
  labs(
    title = "Speed of Light vs. Temperature by Measurement Distinctness",
    x = "Temperature (°F)",
    y = "Velocity in Vacuum (km/s)",
    color = "Distinctness"
  ) +
  theme_minimal()

# Visualization 2: Velocity vs Distinctness (boxplot)
ggplot(df_q2, aes(Distinctness, VelocityVacuum)) +
  geom_boxplot() +
  labs(title = "Speed of Light by Measurement Distinctness")

# Visualization 3: Velocity over Time
ggplot(df_q2, aes(Date, VelocityVacuum)) +
  geom_point() +
  geom_smooth() +
  labs(title = "Speed of Light Measurements Over Time")
```

**Observations**:

-   **Visualization 1: Speed of Light vs. Temperature by Measurement
    Distinctness**
    -   The three straight lines all tilt upward, which means
        Michelson’s measured speed went up a little on warmer days. The
        red line (poor images) rises the fastest and has the most
        spread, so bad image quality shows the strongest temperature
        effect and the noisiest data. The blue line (good images) is
        almost flat and the points cluster tightly, meaning when the
        images were clear his measurements were more consistent and
        barely changed with temperature.

<!-- -->

-    **Visualization 2: Speed of Light by Measurement Distinctness**

    -   As image quality improves from 1 to 3, the middle of each box
        (the median) shifts higher: poor images sit lowest, fair are in
        the middle, and good images are highest. The “good” box is also
        taller, showing more variation in those best trials. Each group
        has a few dots outside the whiskers—those are outliers, like one
        very low poor‐quality reading near 299,700 km/s or a couple of
        high good‐quality readings above 300,100 km/s.

<!-- -->

-    **Visualization 3: Speed of Light Measurements Over Time**

    -   The smooth blue line starts just above 299,945 km/s in early
        June, dips down to near 299,920 km/s around mid‐month, then
        creeps back up toward 299,930 km/s by month’s end. You can also
        see points spread out more in the middle of the month,
        suggesting the experiment was a bit less stable then—maybe due
        to changing conditions or equipment drift.

## Bibliography

-   [1] Michelson, [Experimental Determination of the Velocity of
    Light](https://play.google.com/books/reader?id=343nAAAAMAAJ&hl=en&pg=GBS.PA115)
    (1880) 
-   [2] Henrion and Fischhoff, [Assessing Uncertainty in Physical
    Constants](https://www.cmu.edu/epp/people/faculty/research/Fischoff-Henrion-Assessing%20uncertainty%20in%20physical%20constants.pdf)
    (1986) 
-   [3] BYU video about a [Fizeau-Foucault
    apparatus](https://www.youtube.com/watch?v=Ik5ORaaeaME), similar to
    what Michelson used.
