---
title: "Aluminum Data"
author: "Esther Aduamah"
date: 2025-19-02
output:
  github_document:
    toc: true
prerequisites:
  - e-data03-pivot-basics
---

*Purpose*: When designing structures such as bridges, boats, and planes, the design team needs data about *material properties*. Often when we engineers first learn about material properties through coursework, we talk about abstract ideas and look up values in tables without ever looking at the data that gave rise to published properties. In this challenge you'll study an aluminum alloy dataset: Studying these data will give you a better sense of the challenges underlying published material values.

In this challenge, you will load a real dataset, wrangle it into tidy form, and perform EDA to learn more about the data.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|------------------------|------------------------|------------------------|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**

```{r setup}
library(tidyverse)
```

*Background*: In 1946, scientists at the Bureau of Standards tested a number of Aluminum plates to determine their [elasticity](https://en.wikipedia.org/wiki/Elastic_modulus) and [Poisson's ratio](https://en.wikipedia.org/wiki/Poisson%27s_ratio). These are key quantities used in the design of structural members, such as aircraft skin under [buckling loads](https://en.wikipedia.org/wiki/Buckling). These scientists tested plats of various thicknesses, and at different angles with respect to the [rolling](https://en.wikipedia.org/wiki/Rolling_(metalworking)) direction.

# Loading and Wrangle

<!-- -------------------------------------------------- -->

The `readr` package in the Tidyverse contains functions to load data form many sources. The `read_csv()` function will help us load the data for this challenge.

```{r load}
## NOTE: If you extracted all challenges to the same location,
## you shouldn't have to change this filename
filename <- "./data/stang.csv"

## Load the data
df_stang <- read_csv(filename)
df_stang

```

Note that these data are not tidy! The data in this form are convenient for reporting in a table, but are not ideal for analysis.

### **q1** Tidy `df_stang` to produce `df_stang_long`. You should have column names `thick, alloy, angle, E, nu`. Make sure the `angle` variable is of correct type. Filter out any invalid values.

*Hint*: You can reshape in one `pivot` using the `".value"` special value for `names_to`.

```{r q1-task}
## TASK: Tidy `df_stang`

df_stang_long <- df_stang %>%
  pivot_longer(
    cols = -c(thick, alloy),  # Pivot all except these columns
    names_to = c(".value", "angle"),  # Special .value token
    names_sep = "_"  # Split names at underscore
  ) %>%
  mutate(angle = as.integer(angle)) %>%  # Convert angle to integer
  filter(E > 0 & nu > 0)  # Remove invalid values
```

Use the following tests to check your work.

```{r q1-tests}
## NOTE: No need to change this
## Names
assertthat::assert_that(
              setequal(
                df_stang_long %>% names,
                c("thick", "alloy", "angle", "E", "nu")
              )
            )
## Dimensions
assertthat::assert_that(all(dim(df_stang_long) == c(26, 5)))
## Type
assertthat::assert_that(
              (df_stang_long %>% pull(angle) %>% typeof()) == "integer"
            )

print("Very good!")
```

I cannot figure out this element error for the life of me– I've been too out of it this week to go to office hours with general life things, but I will get it fixed afterwards. It's just brute forced for now and I apologize, but idk what was going on.

# EDA

<!-- -------------------------------------------------- -->

## Initial checks

<!-- ------------------------- -->

### **q2** Perform a basic EDA on the aluminum data *without visualization*. Use your analysis to answer the questions under *observations* below. In addition, add your own *specific* question that you'd like to answer about the data---you'll answer it below in q3.

```{r q2-task}
##
df_stang_long %>%
  summarise(
    min_E = min(E, na.rm = TRUE),
    max_E = max(E, na.rm = TRUE),
    mean_E = mean(E, na.rm = TRUE),
    median_E = median(E, na.rm = TRUE),
    sd_E = sd(E, na.rm = TRUE),
    min_nu = min(nu, na.rm = TRUE),
    max_nu = max(nu, na.rm = TRUE),
    mean_nu = mean(nu, na.rm = TRUE),
    median_nu = median(nu, na.rm = TRUE),
    sd_nu = sd(nu, na.rm = TRUE)
  )

df_stang_long %>%
  summarise_all(~ sum(is.na(.)))

df_stang_long %>%
  summarise(num_alloys = n_distinct(alloy))

unique(df_stang_long$angle)

unique(df_stang_long$thick)
```

**Observations**:

-   Is there "one true value" for the material properties of Aluminum?
    -   No, the properties varies by alloy, angle, and thickness.
-   How many aluminum alloys are in this dataset? How do you know?
    -   The data set contains 1 aluminum alloy as it says num_alloys=1. The summaries num_alloy helps to count it.
-   What angles were tested?
    -   0, 45, and 90 degrees
-   What thicknesses were tested?
    -   0.022, 0.032, 0.064, and 0.081 inches.
-   My question: Does the elastic modulus (`E`) change more with angle or thickness?

## Visualize

<!-- ------------------------- -->

### **q3** Create a visualization to investigate your question from q2 above. Can you find an answer to your question using the dataset? Would you need additional information to answer your question?

```{r q3-task}
## TASK: Investigate your question from q1 here
## wrote about elastic modules and the comparison with the other variables

library(ggplot2)

df_stang_long %>%
  ggplot(aes(x = thick, y = E, color = factor(angle))) +  
  geom_point(size = 3) +  
  geom_smooth(method = "lm", se = FALSE, color = "black") +  # Trend line
  labs(
    title = "Effect of Thickness on Elastic Modulus (E)",
    x = "Thickness (inches)",
    y = "Elastic Modulus (E)",
    color = "Angle"
  ) +
  theme_minimal() 
```

**Observations**:

-   The scatter plot shows that there's a negative relationship between thickness and elastic modulus (E), meaning that thicker aluminum samples tend to have lower stiffness. The trend line backs this up, illustrating a steady drop in E as thickness goes up. On the flip side, it seems that angle (0°, 45°, 90°) doesn't really affect E much, since the values at different angles are quite similar at each thickness level. If the angle were a key factor, we’d expect to see more noticeable groupings or changes in E values, which isn't the case here. So, it seems like thickness has a bigger impact on E than angle does. However, to be sure about this trend and to rule out other factors, we’d need more details—like measurements at intermediate thicknesses, different aluminum alloys, and tests under various conditions.

### **q4** Consider the following statement:

> "A material's property (or material property) is an intensive property of some material, i.e. a physical property that does not depend on the amount of the material."[2]

Note that the "amount of material" would vary with the thickness of a tested plate. Does the following graph support or contradict the claim that "elasticity `E` is an intensive material property." Why or why not? Is this evidence *conclusive* one way or another? Why or why not?

```{r q4-vis}
## NOTE: No need to change; run this chunk
df_stang_long %>%

  ggplot(aes(nu, E, color = as_factor(thick))) +
  geom_point(size = 3) +
  theme_minimal()
```

**Observations**:

-   Does this graph support or contradict the claim above?
    -   The graph throws a wrench in the idea that E is an intensive material property since the elastic modulus (E) changes with thickness. If E were genuinely intensive, it would stay the same no matter how thick the material is, because intensive properties shouldn't rely on the quantity of material. Yet, in this graph, different thicknesses—shown in different colors—reveal various clusters of E values. This variation suggests that elasticity might actually depend on the amount of material being tested.
-   Is this evidence *conclusive* one way or another?
    -   No, this evidence isn’t completely conclusive. While the graph indicates that E changes with thickness, there might be other factors at play, like testing conditions, measurement errors, or slight variations in how the material was processed. Plus, this dataset only covers one specific type of aluminum alloy, so we can’t really make broad conclusions for all materials. To really figure out if E acts like an intensive property across the board, we’d need more data, including different alloys, various thickness measurements, and controlled testing setups.

# References

<!-- -------------------------------------------------- -->

[1] Stang, Greenspan, and Newman, "Poisson's ratio of some structural alloys for large strains" (1946) Journal of Research of the National Bureau of Standards, (pdf link)[<https://nvlpubs.nist.gov/nistpubs/jres/37/jresv37n4p211_A1b.pdf>]

[2] Wikipedia, *List of material properties*, accessed 2020-06-26, (link)[<https://en.wikipedia.org/wiki/List_of_materials_properties>]
