---
title: "The Islands, Part 1: Design"
author: "Esther A"
date: 2025-04-03
output:
  github_document:
    toc: true
prerequisites:
  - e-stat12-randomization
---

*Purpose*: So far in this class, I've handed you each dataset. But much of the important work of statistics happens *before* data collection. Issues with data collection generally can't be fixed by fancy analysis after the fact. Therefore, in this *two part* challenge, you will first *plan* and then *execute* a statistical project.

To do this, we'll make use of The Islands, an extremely detailed simulation of a virtual population, developed by researchers at the University of Queensland. This is a fascinating sandbox that helps illustrate the difficulties and complexities of collecting real data, while still being more tractable than collecting data from real humans.

This is part 1 of 2. In this part you will *plan* your statistical project, particularly your data collection. Sound data collection is called *statistical design*.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|------------------------|------------------------|------------------------|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**

# Setup

<!-- ----------------------------------------------------------------------- -->

```{r setup}
library(tidyverse)
library(rsample)

filename_random <- "./data/helvig-random.csv"
```

### **q1** Orientation

Log into [The Islands](https://islands.smp.uq.edu.au/index.php) and head to the [Visitor Center](https://islands.smp.uq.edu.au/visitors.php) on Providence island (to the east). Watch the first three videos on the `Guides` tab.

*Note:* There is no deliverable for this task. BUT, you'll need to be oriented to The Islands in order to do *any* of the tasks below.

The Islands is an *incredibly* detailed simulation of a real population. Islanders are born, die, and move around. The Islands was designed as a teaching tool to help students learn how to apply statistics to collecting data in the real world.

### **q2** Occupied homes

Find the total number of homes in Helvig, and count the number of unoccupied homes. Answer the questions below.

NB. When Zach counted the homes (on 2025-01-08), he found 20 unoccupied out of 536 total. That means the occupied percentage was about 96%. But you still need to count, because your numbers *will* be different!

**Observations**:

-   What is the total number of homes in Helvig?
    -   547
-   What is the number of unoccupied homes? (*Hint*: This is not given anywhere. You will have to count them!)
    -   20 houses
-   What percent of homes are *occupied*?
    -   96% occupied (527 out of 547 houses)
-   Are there any sources of *real* uncertainty in the percent occupied you calculated?
    -   When people move in and out over time, the occupancy status may change, but our count
-   Are there any sources of *erroneous* uncertainty in the percent occupied you calculated?
    -   Something would've happened where not every house was checked because the person checking didn't go to every house.

Zach looked at the first 25 homes in Helvig and recorded the `age` and `name` of every person in those homes. These people are provided in `helvig-seq.csv`.

```{r}
## NOTE: Do not edit this
df_sample_seq <- read_csv("./data/helvig-seq.csv")
df_sample_seq 
```

You'll use this dataset as a starting point to figure out prevalent *last names* in Helvig.

### **q3** Find common names: Sequential sample

Complete the code below to find the prevalence of the most common *last name* in Helvig using the sample `df_sample_seq`. Answer the questions below.

Make sure to include in `df_q3` the counts as a column named `n`, and the prevalence (number of occurrence divided by total people) as a column `p`. Also, make sure to sort the data in *descending* order of prevalence.

*Hint*: You will have to use what you've learned about string handling to extract the *last names* only!

```{r}
library(dplyr)

## TASK: Compute the prevalence and sort
#df_q3 <- df_sample_seq %>% 
  #mutate(LastName = word(name, 2)) %>% 

## TODO: Complete this code

  #count(LastName, sort=True)

# df_q3 <- df_sample_seq %>% 
#   names(table(df$names))[which.max(table(df$words))]
# 
# df_q3

df_q3 <- df_sample_seq %>%
  mutate(last_name = word(name, -1)) %>%
  count(last_name, name = "n") %>%
  mutate(p = n / sum(n)) %>%
  arrange(desc(p))

df_q3
```

Use the following to check your work.

```{r q3-tests}
## NOTE: No need to change this
## Check that data has `p` column and is in descending order
assertthat::assert_that(
              all(df_q3 %>%
                 mutate(d = p - lead(p)) %>% 
                 filter(!is.na(d)) %>% 
                 pull(d) >= 0
              )
            )
print("Very good!")
```

*Observations*

-   What last name is most prevalent in `df_sample_seq`?
    -   Sorensen is the most prevalent in the df_sample_seq with eight occurrences.
-   Is this sample representative of *all* houses in Helvig? Why or why not?
    -   No, this is merely a selection of Helvig's first twenty-five homes. The predominance of a last name may be distorted by a large household in the first 25 homes.

In the exercises, we talked about the importance of random sampling. In the previous challenge, we were able to *simulate* a random sample by running a few lines of code. But in the real world, we have to work harder to gather a random sample. We'll do this in two stages: *plan* then *collect*.

### **q4** Plan a random sample

Complete the code below to draw a sample of size `n=25`. Replace `n_houses` with the (current) total number in Helvig.

```{r}
## TASK: Set the parameters for this code block

## Select a random sample of houses
n_houses <- 547 # Total number of houses
n_sample <- 25  # Desired sample size

set.seed(101)   # Set a seed for reproducibility

df_numbers_random <- 
  tibble(
    house = sample(
        1:n_houses,     # All integers from 1 to n_houses
        n_sample,       # Size of our sample
        replace = FALSE # Sample *WITHOUT* replacement
      )
  ) %>% 
  # Arrange for our data collection convenience
  arrange(house)

# Pull the column so we can list just the house numbers
df_numbers_random %>% 
  pull(house)
```

Use the following code to check your results.

```{r}
## NOTE: No need to change this
assertthat::assert_that(
  all(dim(df_numbers_random) == c(25, 1))
)
```

### **q5** Collect the random sample

Gather the names of all people in the homes you listed in the previous exercise. Match the same columns as `df_sample_seq`; those are, `house`, `age`, `name`. Make sure to include `NA` rows for houses with no occupants. Save your data as a CSV with the filename provided in the variable `filename_random`. Answer the questions below.

```{r}
## NOTE: Do not edit
filename_random
```

Note that this points to the `data/` subdirectory in your `challenges` folder.

The following code will load your data.

```{r}
## NOTE: Do not edit
df_sample_random <- 
  read_csv(filename_random)
```

Use the following to check your work.

```{r q5-tests}
## NOTE: No need to change this
# Check that the dataset has the correct column names
assertthat::assert_that(setequal(
  df_sample_random %>% names(),
  df_sample_seq %>% names()
))

# Check that all of the house numbers in the dataset match those that were planned
numVsamp <- 
  anti_join(
    df_numbers_random,
    df_sample_random %>% distinct(house),
    by = "house"
  ) %>% 
  pull(house)
assertthat::assert_that(
  length(numVsamp) == 0,
  msg = str_c("You are missing the houses: ", str_c(numVsamp, collapse = ", "))
)

sampVnum <- 
  anti_join(
    df_sample_random %>% distinct(house),
    df_numbers_random,
    by = "house"
  ) %>% 
  pull(house)
assertthat::assert_that(
  length(sampVnum) == 0,
  msg = str_c("You have extra houses: ", str_c(sampVnum, collapse = ", "))
)

print("Great work!")
```

*Observations*

-   Which sample---sequential or random---is more *representative* of all homes Helvig? Why?
    -   Random
    -   The sequential sample only looks at the first 25 houses, which may cluster families together (since relatives tend to live near each other). The random sample spreads observations across the entire town, better representing the true name distribution.

### **q6** Find common names: Random sample

Run the code below to find the prevalence of the most common *last name* in Helvig using the sample `df_sample_random`. Answer the questions below.

```{r}
# NOTE: No need to edit; run and answer the questions below
df_sample_random %>% 
  mutate(last = str_extract(name, "\\w+$")) %>% 
  count(last) %>% 
  arrange(desc(n)) %>% 
  mutate(p = n / sum(n))
```

*Observations*

-   Did you find any highly prevalent names using `df_sample_random` that you *didn't* find in q3 (using `df_sample_seq`)? Write them here.
    -   Collins and Morris appear a bunch of times. Names like Kono, Ludwig, Nowak, Pillai, Ramanujan, Svendsen, and Wilson appeared in the random sample but weren't in the sequential sample's top names - The sequential sample was dominated by Sorensen (8 occurrences), while the random sample shows more diversity.
-   Is there any reason that people with the same last name might tend to *live near each other*?
    -   Maybe people from the same country or region fo the world would live by each other. They could not just be from the same country, but relatives as family often lives next to each other.

You should have found some difference between the sequential and random samples. This is because we're only working with a *sample*---a limited number of observations from the entire population. We could go to every single house in Helvig and determine *exactly* how many people of each name there are (that's called a *census*), but the point of statistical inference is that we can make statements about a population using only a sample. What makes statistical inference powerful is that we can determine *how confident* we should be in our results, based on the size of our sample.

To do this, we'll use the bootstrap approach that you saw in the exercise sequence. We'll start by building a helper function.

### **q7** Write a helper function

Complete the code below to write a helper function. Your function will compute the proportion of people in a sample that have a user-specified last name.

```{r}
## TASK: Write a helper function that takes a dataframe with full names 
#  (provided in a `name` column), removes any invalid rows, and computes the
#  proportion of individuals with the user-specified `last` name (returned
#  in an `estimate` column).
name_prevalence <- function(df, last = "Collins") {
  df %>% 
    filter(!is.na(name)) %>%               # Remove empty rows
    mutate(last_name = word(name, -1)) %>% # Extract last word as surname
    summarise(
      term = "prevalence",                 # Label for the output
      estimate = mean(last_name == last)    # Calculate proportion
    )
}
```

Use the following code to check your results.

```{r}
## NOTE: No need to change this
# Find the most prevalent name in the data
last_most <- 
  df_sample_random %>% 
  mutate(last = str_extract(name, "\\w+$")) %>% 
  count(last) %>% 
  arrange(desc(n)) %>% 
  slice(1) %>% 
  pull(last)

# Ensure correct columns
assertthat::assert_that(
  setequal(
    tibble(name = c("James")) %>% name_prevalence(., last = "James") %>% names(),
    c("term", "estimate")
  ),
  msg = "Your code should result a dataframe with just two columns: `term` and `estimate`"
)

# Ensure NA handling
assertthat::assert_that(
  !(tibble(name = c(NA_character_, "James")) %>% 
    name_prevalence(., last = "James") %>% 
    pull(estimate) %>% 
    is.na()),
  msg = "Ensure your code properly ignores NA's"
)

# Check for correctness
assertthat::assert_that(
  name_prevalence(df_sample_random, last = last_most) %>% pull(estimate) ==
    mean(str_detect(df_sample_random$name, last_most), na.rm = TRUE),
  msg = "Your code computed the wrong value"
)

print("Nice!")
```

### **q8** Construct a bootstrap confidence interval

Choose a prevalent name that you found in q6. Use bootstrap resampling with your helper function to construct a confidence interval for the prevalence of that name. Answer the questions below.

*Hint*: We learned how to do resampling-based inference in `e-stat09`.

```{r}
# TASK: Complete the code below to compute a bootstrap-based confidence interval
df_interval_bootstrap <- 
  df_sample_random %>% 
  bootstraps(., times = 1000) %>% 
  mutate(
    estimate = map(
      splits,
      function(split_df) {
         name_prevalence(analysis(split_df), last = "Collins")
## TODO: Finish this code, using the name_prevalence() helper you implemented
## HINT: Remember that you need to use analysis() when operating on split_df
      }
    )
  ) %>% 
  ## NOTE: No need to edit this line; this uses your bootstrap sample to compute
  # a confidence `int`erval using the percentile method
  int_pctl(., estimate)

df_interval_bootstrap 
```

**Observations**:

-   What is highest possible prevalence for your chosen name, based on the confidence interval you constructed?
    -   21.57%
    -   This means we can be 95% confident the true population prevalence is less than or equal to 21.57%
-   Note that we used the *random* sample with the bootstrap procedure in this task. Could we use the bootstrap to make a confidence interval using the sequential sample (`df_sample_seq`) that would be representative of all of Helvig? Why or why not?
    -   No
    -   The sequential sample's results would be unreliable because: 1. **Geographic clustering**: Families live near each other, artificially inflating name concentrations 2. **Underestimated variability**: Wouldn't account for neighborhood differences 3. **Bias**: First 25 houses might be in a single "family district"

### **q9** Discover possible measurements

Click on a single islander (you can do this from one of the houses), and take a look at what info they can provide. Write down three pieces of info that you're particularly interested in; this will inform your quantity of interest for the final task.

*Hint*: The videos from the [Visitor Center](https://islands.smp.uq.edu.au/visitors.php) will be especially helpful for getting some ideas.

-   Ruler drop test reaction time (cm)
-   Blood alcohol content (BAC)
-   Balance test duration (seconds)

### **q10** Planning a study (TEAMWORK)

Challenge 10 will be a team assignment where you plan and execute a statistical study using The Islands. You should meet with your learning team for Challenge 10 to complete the steps below---though everyone is responsible for including the plan in their individual Challenge 08 submission.

#### Population

-   (What population are you going to study?)
    -   All of the Islanders? Only the residents of Helvig? Participants meeting certain criteria (in which case, you need to list *covariates*)?
    -   We will be studying the Akkeshi Population in The Islands. We used random sampling to get a random sampling of 30 houses, with the participants being over 18 years old.

#### Quantity of interest

-   (Write your quantity of interest here)
-   Our quantity of interest is the change reaction time measured by the length in centimeters caught after the ruler is dropped

#### Covariates

-   (Write any relevant covariates here)
-   (There's probably more than one!)
    -   Weight, height, age, amount of alcohol consumed, reaction time before consumption, reaction time after consumption 

#### Observation or experiment?

The Islands allows you to ask islanders to complete tasks. If you just take measurements on your participants, then it's an *observational study*. But if you also introduce something that's meant to change the outcome of a measurement (e.g., drinking coffee before taking a test), that's called an *experimental study*. You need to decide whether your study is observational or experimental.

-   (Observation or experiment?)
    -   Experiment because we are introducing alcohol to the residents of Akkeshi and recording the changes in reaction times.

#### Question / Hypothesis

-   (Write your question / hypothesis here)
    -   This could be an observational question, like "How are X and Y related in the population?"
    -   This could be an experimental question, like "What effect does X have on Y in the population?"
    -   How do the height and weight of an individual affect their reaction time when drinking 4 shots of alcohol? We changed it from our original question to include height and weight.

#### Sampling plan

-   (What steps will you take to collect the data?)
    -   First, we all met together at the same time to create/open our The Islands logins together. As a team, we decided to focus on Akkeshi. We did a random sampling with set.seed to figure out which houses we were going to be checking. We put all of these numbers in a spreadsheet. We split up the households so every member of the team would have five houses to cover; this was so we did our test relative in the same time period in Akkeshi. We recorded every member of the household’s height, weight, and name as long as they were 18 and consented. If they did not consent we entered them as “NA” on our data sheet. We all recorded the Blood Alcohol Content (B.A.C.) and reaction time (by doing the Ruler Test) before and after alcohol consumption for every single person. We gave everyone four tequila shots back to back and waited five real-world minutes before checking their blood alcohol content. This is so the alcohol could sit for a bit and get into their systems. We then re-recorded their reaction time and went on to the next person in the household, or new house if no one else fit the requirements for our study.
-   (How will you ensure the data is representative of your chosen population?)
    -   As we did a random sampling, the data is representative as we have thirty houses from nine to 447 all on Akkeshi Island.
-   (For experiments only: How will you ensure any effects you observe are due to the treatment, and not due to other factors?)
    -   We all did our experiments at the same time and waited the same amount of time (5 minutes), making sure that time did not affect us. We also collected all of our data together
-   (How will you choose your sample size?)
    -   This should be a mix of how confident you want to be in your results, and what is actually *feasible* for your research team.
        -   30 houses because it was manageable for us to collect the data and we considered that in some houses there is more than one person and not every person will give us their consent to participate in our experiment.
